---
title: "4-LineGraphs"
format: gfm
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(paged.print = FALSE)
```

```{r message=FALSE}
library(tidyverse)
library(gcookbook)
library(patchwork)
```

[Source](https://r-graphics.org/RECIPE-LINE-GRAPH-LINE-APPEARANCE.html)

# Line graphs

## Basic

```{r}
ggplot(BOD, aes(x = Time, y = demand)) +
  geom_line()
```

> A continuous (numeric) variable can be treated as a discrete (categorical) one. When x is a factor, must use `group = 1`

```{r}
BOD1 <- BOD
BOD1$Time <- factor(BOD1$Time)

ggplot(BOD1, aes(x = Time, y = demand, group = 1)) +
  geom_line()
```

> Setting range with `ylim()` or `expand_limits()`

```{r}
p1 <- BOD |> 
  ggplot(aes(x = Time, y = demand)) +
  geom_line() +
  ylim(0, max(BOD$demand))
p2 <- BOD |> 
  ggplot(aes(x = Time, y = demand)) +
  geom_line() +
  expand_limits(y = 0)
p1 + p2
```

## Adding points to lines

```{r}
p1 <- ggplot(worldpop, aes(x = Year, y = Population)) +
  geom_line() +
  geom_point()
p2 <- ggplot(worldpop, aes(x = Year, y = Population)) +
  geom_line() +
  geom_point() +
  scale_y_log10()
p1 + p2
```

## Multiple lines

```{r}
ggplot(tg, aes(x = dose, y = length, colour = supp, linetype = supp)) +
  geom_line()
```

> With categorical variable must use `group`

```{r}
ggplot(tg, aes(x = factor(dose), y = length, colour = supp, group = supp)) +
  geom_line()
```

```{r}
p1 <- ggplot(tg, aes(x = dose, y = length, shape = supp)) +
  geom_line() +
  geom_point(size = 4)
p2 <- ggplot(tg, aes(x = dose, y = length, fill = supp)) +
  geom_line() +
  geom_point(size = 4, shape = 21)
p1 + p2
```

> Adjust for overlapping points

```{r}
ggplot(tg, aes(x = dose, y = length, shape = supp)) +
  geom_line(position = position_dodge(0.2)) +
  geom_point(size = 4, position = position_dodge(0.2))
```

## Line appearance

```{r}
ggplot(BOD, aes(x = Time, y = demand)) +
  geom_line(linetype = "dashed", size = 1, colour = "blue")
```

```{r}
ggplot(tg, aes(x = dose, y = length, colour = supp)) +
  geom_line() +
  scale_colour_brewer(palette = "Set1")
```

> Grouping must be specified if both lines have the same property

```{r}
p1 <- ggplot(tg, aes(x = dose, y = length, group = supp)) +
  geom_line(colour = "darkgreen", size = 1.5)

# This will be automatically grouped by color
p2 <- ggplot(tg, aes(x = dose, y = length, colour = supp)) +
  geom_line(linetype = "dashed") +
  geom_point(shape = 22, size = 3, fill = "white")
p1 + p2
```

## Point appearance

```{r}
ggplot(BOD, aes(x = Time, y = demand)) +
  geom_line() +
  geom_point(size = 4, shape = 22, colour = "darkred", fill = "pink")
```

> Multi-line

```{r}
ggplot(tg, aes(x = dose, y = length, fill = supp)) +
  geom_line(position = position_dodge(0.2)) +
  geom_point(shape = 21, size = 3, position = position_dodge(0.2)) +
  scale_fill_manual(values = c("black","white"))
```

## Shaded area

```{r}
sunspot_year <- data.frame(
  Year = as.numeric(time(sunspot.year)),
  Sunspots = as.numeric(sunspot.year)
)

p1 <- ggplot(sunspot_year, aes(x = Year, y = Sunspots)) +
  geom_area(color = "black", fill = "blue", alpha = 0.2)

p2 <- ggplot(sunspot_year, aes(x = Year, y = Sunspots)) +
  geom_area(fill = "blue", alpha = .2) +
  geom_line()

p1 + p2
```

## Stacked area graph

> map a factor to `fill`

```{r}
ggplot(uspopage, aes(x = Year, y = Thousands, fill = AgeGroup)) +
  geom_area()
```

> `ggplot` requires data to be in long format, so may need to be converted from wide

```{r}
ggplot(uspopage, aes(x = Year, y = Thousands, fill = AgeGroup)) +
  geom_area(colour = "black", size = .2, alpha = .4) +
  scale_fill_brewer(palette = "Blues")
```

```{r}
ggplot(uspopage, aes(x = Year, y = Thousands, 
                     fill = AgeGroup, order = dplyr::desc(AgeGroup))) +
  geom_area(colour = NA, alpha = .4) +
  scale_fill_brewer(palette = "Blues") +
  geom_line(position = "stack", size = .2)
```

## Proportional stacked area graph

> use `position = "fill"`

```{r}
p1 <- ggplot(uspopage, aes(x = Year, y = Thousands, fill = AgeGroup)) +
  geom_area(position = "fill", colour = "black", size = .2, alpha = .4) +
  scale_fill_brewer(palette = "Blues")

# labels as percent
p2 <- ggplot(uspopage, aes(x = Year, y = Thousands, fill = AgeGroup)) +
  geom_area(position = "fill", colour = "black", size = .2, alpha = .4) +
  scale_fill_brewer(palette = "Blues") +
  scale_y_continuous(labels = scales::percent)

p1 + p2
```

## Adding a confidence region

> `geom_ribbon()...` place before `geom_line()`

```{r}
climate_mod <- climate |> 
  filter(Source == "Berkeley") |> 
  select(Year, Anomaly10y, Unc10y)
climate_mod |> head()
```

```{r}
ggplot(climate_mod, aes(x = Year, y = Anomaly10y)) +
  geom_ribbon(aes(ymin = Anomaly10y - Unc10y, ymax = Anomaly10y + Unc10y), alpha = 0.2) +
  geom_line()
```

> Dotted lines instead of shaded regions

```{r}
ggplot(climate_mod, aes(x = Year, y = Anomaly10y)) +
  geom_line(aes(y = Anomaly10y - Unc10y), colour = "grey50", linetype = "dotted") +
  geom_line(aes(y = Anomaly10y + Unc10y), colour = "grey50", linetype = "dotted") +
  geom_line()
```
