---
title: "7 - Annotations"
format: gfm
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(paged.print = FALSE)
```

```{r message=FALSE}
library(tidyverse)
library(gcookbook)
library(patchwork)
```

[Source](https://r-graphics.org/RECIPE-LINE-GRAPH-LINE-APPEARANCE.html)

# Annotations

## Text

```{r}
p <- ggplot(faithful, aes(x = eruptions, y = waiting)) +
  geom_point()

p +
  annotate("text", x = 3, y = 48, label = "Group 1") +
  annotate("text", x = 4.5, y = 66, label = "Group 2")
```

> With font formatting

```{r}
p +
  annotate("text", x = 3, y = 48, label = "Group 1",
           family = "serif", fontface = "italic", 
           colour = "darkred", size = 3) +
  annotate("text", x = 4.5, y = 66, label = "Group 2",
           family = "serif", fontface = "italic", 
           colour = "darkred", size = 3)
```

> for continuous axes, use `Inf` and `-Inf` for the edges. Calculated values can be used for placement.

```{r}
p +
  annotate("text", x = -Inf, y = Inf, 
           label = "Upper left", 
           hjust = -.2, vjust = 2) +
  annotate("text", 
           x = mean(range(faithful$eruptions)), 
           y = -Inf, vjust = -0.4,
           label = "Bottom middle")
```

## Annotations with mathematical expression

> Reference `?plotmath`

```{r}
p <- ggplot(data.frame(x = c(-3,3)), aes(x = x)) +
  stat_function(fun = dnorm)

p + 
  annotate("text", x = 2, y = 0.3, parse = TRUE,
           label = "frac(1, sqrt(2 * pi)) * e ^ {-x^2/2}")
```

```{r}
p +
  annotate("text", x = 0, y = 0.05, parse = TRUE, size = 4,
           label = "'Function:  ' * y==frac(1, sqrt(2*pi)) * e^{-x^2/2}")
```

## Adding lines

> `geom_hline()`, `geom_vline()`, `geom_abline()`

```{r}
hw_plot <- ggplot(heightweight, aes(x = ageYear, y = heightIn, colour = sex)) +
  geom_point()

hw_plot +
  geom_hline(yintercept = 60) +
  geom_vline(xintercept = 14)
```

```{r}
hw_plot +
  geom_abline(intercept = 37.4, slope = 1.75)
```

> draw lines from data frames

```{r}
hw_means <- heightweight |> 
  group_by(sex) |> 
  summarise(heightIn = mean(heightIn))
hw_means
```

```{r}
hw_plot +
  geom_hline(
    data = hw_means,
    aes(yintercept = heightIn, colour = sex),
    linetype = "dashed",
    linewidth = 1
  )
```

> For discrete variables, levels are numbered or access by name with `which(levels))`

```{r}
pg_plot <- ggplot(PlantGrowth, aes(x = group, y = weight)) +
  geom_point()
```

```{r}
pg_plot + 
  geom_vline(xintercept = 2)
```

```{r}
pg_plot +
  geom_vline(xintercept = which(levels(PlantGrowth$group) == "ctrl"))
```

> Line segments and arrows
>
> \`annotate("segment")

```{r}
p <- ggplot(filter(climate, Source == "Berkeley"), aes(x = Year, y = Anomaly10y)) +
  geom_line()
```

```{r}
p +
  annotate("segment", x = 1950, xend = 1980, y = -.25, yend = -.25)
```

> Adding arrowheads or segment end lines

```{r}
library(grid)

p +
  annotate("segment", x = 1850, xend = 1820, y = -.8, yend = -.95,
           color = "blue", linewidth = 2, arrow = arrow()) +
  annotate("segment", x = 1950, xend = 1980, y = -.25, yend = -.25,
           arrow = arrow(ends = "both", angle = 90, length = unit(.2, "cm")))
```

## Shaded rectangles

> `annotate("rect")`

```{r}
p <- ggplot(filter(climate, Source == "Berkeley"), aes(x = Year, y = Anomaly10y)) +
  geom_line()

p +
  annotate("rect", xmin = 1950, xmax = 1980, ymin = -1, ymax = 1,
           alpha = .1, fill = "blue")
```

## Highlighting items

```{r}
pg_mod <- PlantGrowth |> 
  mutate(hl = recode(group, "ctrl" = "no", "trt1" = "no", "trt2" = "yes"))
```

```{r}
ggplot(pg_mod, aes(x = group, y = weight, fill = hl)) +
  geom_boxplot() +
  scale_fill_manual(values = c("gray85", "#ffddcc"), guide = "none")
```

or, since there are only three

```{r}
ggplot(pg_mod, aes(x = group, y = weight, fill = group)) +
  geom_boxplot() +
  scale_fill_manual(values = c("gray85", "gray85", "#ffddcc"), guide = "none")
```

## Error Bars

> `geom_errorbar()`

```{r}
ce_mod <- cabbage_exp |> 
  filter(Cultivar == "c39")

p <- ggplot(ce_mod, aes(x = Date, y = Weight))
```

```{r}
p +
  geom_col(fill = "white", color = "black") +
  geom_errorbar(aes(ymin = Weight - se, ymax = Weight + se), width = .2)
```

```{r}
p + 
  geom_line(aes(group = 1)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = Weight - se, ymax = Weight + se), width = .2)
```

> For a bar graph with groups of bars, the error bars must also be *dodged*; otherwise, they’ll have the exact same *x* coordinate and won’t line up with the bars.

```{r}
# Bad
ggplot(cabbage_exp, aes(x = Date, y = Weight, fill = Cultivar)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = Weight - se, ymax = Weight + se),
                position = "dodge", width = .2)
```

```{r}
# Good
ggplot(cabbage_exp, aes(x = Date, y = Weight, fill = Cultivar)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = Weight - se, ymax = Weight + se),
                position = position_dodge(0.9), width = .2)
```

```{r}
ggplot(cabbage_exp, aes(x = Date, y = Weight, fill = Cultivar)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = Weight - se, ymax = Weight + se),
                position = position_dodge(0.9), width = .2)
```

> Draw error bars first if they are a different color than the lines and points
>
> Make sure to dodge all geometric elements

```{r}
pd <- position_dodge(.3)

ggplot(cabbage_exp, aes(x = Date, y = Weight, colour = Cultivar, group = Cultivar)) +
  geom_errorbar(
    aes(ymin = Weight - se, ymax = Weight + se),
    width = .2, 
    linewidth = 0.25, 
    color = "black",
    position = pd) +
  geom_line(position = pd) +
  geom_point(position = pd)
```

```{r}
ggplot(cabbage_exp, aes(x = Date, y = Weight, colour = Cultivar, group = Cultivar)) +
  geom_errorbar(
    aes(ymin = Weight - se, ymax = Weight + se),
    width = .2,
    size = 0.25,
    colour = "black",
    position = pd
  ) +
  geom_line(position = pd) +
  geom_point(position = pd, size = 2.5)
```

## Annotate individual facets

```{r}
mpg_plot <- ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(. ~ drv)

f_labels <- data.frame(drv = c("4", "f", "r"),
                       label = c("4wd", "Front", "Rear"))

mpg_plot +
  geom_text(x = 6, y = 40, aes(label = label), data = f_labels)
```

```{r}
lm_labels <- function(dat) {
  mod <- lm(hwy ~ displ, data = dat)
  formula <- sprintf("italic(y) == %.2f %+.2f * italic(x)",
                     round(coef(mod)[1], 2), round(coef(mod)[2], 2))
  r <- cor(dat$displ, dat$hwy)
  r2 <- sprintf("italic(R^2) == %.2f", r^2)
  data.frame(formula = formula, r2 = r2, stringsAsFactors = FALSE)
}
```

```{r}
labels <- mpg |> 
  group_by(drv) |> 
  do(lm_labels(.))
labels
```

```{r}
mpg_plot +
  geom_smooth(method = lm, se = FALSE) +
  geom_text(data = labels, aes(label = formula), 
            x = 3, y = 40, parse = TRUE, hjust = 0)
```

> To just display values

```{r}
labels <- mpg |> 
  group_by(drv) |> 
  summarise(r2 = cor(displ, hwy)^2)

labels$r2 <- sprintf("italic(R^2) == %.2f", labels$r2)

mpg_plot +
  geom_smooth(method = lm, se = FALSE) +
  geom_text(data = labels, aes(label = r2), 
            x = 3, y = 40, parse = TRUE, hjust = 0)
```
